---
title: "GSEA tutorial"
author: "Likai"
date: '2022-07-29'
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r}
setwd('/home/big/tanlikai/Lung/')

```


```{r setup, include=FALSE}
library(Seurat)
library(ggplot2)
library(purrr)
library(dplyr)
library(ggplot2)
library(cowplot)
library(magrittr)

source('/home/big/tanlikai/script/rscripts/funcs.r')
```


```{r gsea_related_packages} 

library(clusterProfiler)
library(msigdbr)
library(enrichplot)
```

## R Markdown
# to generate gene list
# replace GDTcell with your own object, '0' and '1' with the clusters you want to compare

```{r Seuratproject}
GDTcell <- readRDS('public/GDTcell_satija.rds')
```


```{r genelist}
CP01 <- entrezlist_generator(GDTcell, '0', '1')
head(CP01)
```

# genereate reference

# here means I choose homo sapiens, and category 7 (immunology) in Msigdb
# http://www.gsea-msigdb.org/gsea/msigdb
```{r GSEA references}
Mc7 <- msigdbr::msigdbr(species = "Homo sapiens", category = "C7") %>%
  dplyr::select(gs_name, entrez_gene)

# HALLMARK
HALLMARK <-  msigdbr::msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gs_name, entrez_gene)

# KEGG
KEGG  <-   msigdbr::msigdbr(species = "Homo sapiens", category = "C2",subcategory = 'CP:KEGG') %>%
  dplyr::select(gs_name, entrez_gene)
```

```{r}
print(Mc7)
```
# Run GSEA
# TERM2GENE is the gene list
# for mouse project, you need change the OrgDb to org.Mm.eg.db
```{r runGESA}
GSEACP01 <- GSEA(geneList = CP01, TERM2GENE=Mc7,
                 pvalueCutoff = 0.05, pAdjustMethod = "BH") %>% 
  setReadable(OrgDb = org.Hs.eg.db, keyType="ENTREZID")
```
#the result table is in object@result
```{r}
head(GSEACP01@result)

```
#export result as xlsx file 
```{r}
openxlsx::write.xlsx(GSEACP01@result, 'Lung_c0_c1_satijadata_GSEA_c7.xlsx')
```

#Visualization
#select the ID you want to visualize to generate GSEA plot


```{r}
gseaplot2(GSEACP01, geneSetID = 'GSE25087_TREG_VS_TCONV_ADULT_UP')

```


```{r}
gseaplot2(GSEACP01, geneSetID = c('GSE25087_TREG_VS_TCONV_ADULT_UP','GSE26495_NAIVE_VS_PD1LOW_CD8_TCELL_DN'  ))

```
```











